---
title: "Type 2 Diabetes in Pima Native Americans"
author: "Brian Burrows, Eric Sellew, Yonas Shiferaw, and Kelly Yang"
date: "12/3/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Load Packages
library(mosaic)
library(ggplot2)
library(tidyverse)
library(pROC)
library(rpart)
library(partykit)
```

## R Markdown

```{r}
pima <- read.csv("https://pmatheson.people.amherst.edu/Pima.dat", header=FALSE) #Loading data
colnames(pima) <- c("PRG", "PLASMA", "BP", "THICK", "INSULIN", "BODY", "PEDIGREE", "AGE", "RESPONSE") #renaming column names
```


## Prevalence of Diabetes in the Dataset

```{r}
tally(pima$RESPONSE)
268/(500+268)
#Approximately 35% of the individual females in this sample have diabetes.
```


```{r}
<<<<<<< HEAD
filteredpima <- filter(pima,PLASMA>0,BP>0,THICK>0,BODY>0,INSULIN>0)
#We noticed biological impossibilities in the data, such as blood pressures of 0. Since it is not possible to distinguish true zeros from non-entries, we removed all entries that had zeros in PLASMA, BP, THICK, BODY, and INSULIN.
```


```{r}
tally(filteredpima$RESPONSE)
262/(262+130)
#Approximately 67% of the individual females in the filtered dataset have diabetes.
=======
filteredpima <- filter(pima,PLASMA>0,BP>0,THICK>0,BODY>0,INSULIN>0) #creates a dataset that does not have missing values that are denoted by 0
```

```{r}
#Create a stepwise model
model <- glm(RESPONSE~., data=pima, family=binomial) %>%
  MASS::stepAIC(trace=FALSE)
summary(model)

#Stepwise model using the filtered dataset
filteredmodel <- glm(RESPONSE~., data=filteredpima, family=binomial) %>%
  MASS::stepAIC(trace=FALSE)
summary(filteredmodel)
>>>>>>> b2096195ddaa531e7ee6b6a130de116f0e038a1f
```


## EDA

```{r}
summary(filteredpima)
histogram(~PRG, data=filteredpima) #PRG is skewed to the right.
histogram(~PLASMA, data=filteredpima) #Follows a somewhat normal distribution.
histogram(~BP, data=filteredpima) #Also somewhat bell-shaped.
histogram(~THICK, data=filteredpima) #Mostly normal distribution.
histogram(~INSULIN, data=filteredpima) #Skewed to the right.
histogram(~BODY, data=filteredpima) #Somewhat normal distribution, but is slightly skewed to the right.
histogram(~PEDIGREE, data=filteredpima) #Skewed to the right.
histogram(~AGE, data=filteredpima) #Skewed to the right.
```


## Plots

```{r}
#Plot regarding blood pressure
pima_bp <- pima %>%
  mutate(RESPONSE = ifelse(RESPONSE == 0, "Non-Diabetic", "Diabetic"))%>%
  filter(BP > 0) %>%
  group_by(BP, RESPONSE) %>%
  summarize(response_total = n())
ggplot(pima_bp, aes(x=BP, fill=factor(RESPONSE))) +
  geom_density(alpha=0.5) +
  ylab("Density") +
  xlab("Diastolic Blood Pressure") +
  ggtitle("Relationship between Diabetic and Non-Diabetic Groups and Blood Pressure") +
  scale_fill_manual("Diabetic Group", values=c("Red","Blue"))
```


```{r}
#Plot regarding skinfold thickness
pima_thick <- pima %>%
  mutate(RESPONSE = ifelse(RESPONSE == 0, "Non-Diabetic", "Diabetic"))%>%
  filter(THICK > 0) %>%
  group_by(THICK, RESPONSE) %>%
  summarize(response_total = n())
ggplot(pima_thick, aes(x=THICK, fill=factor(RESPONSE))) +
  geom_density(alpha=0.5) +
  ylab("Density") +
  xlab("Triceps Skin Fold Thickness") +
  ggtitle("Relationship between Diabetic and Non-Diabetic Groups and Triceps Skin Fold Thickness") +
  scale_fill_manual("Diabetic Group", values=c("Red","Blue"))
```


```{r}
#Plot regarding plasma level
pima_plasma <- pima %>%
  mutate(RESPONSE = ifelse(RESPONSE == 0, "Non-Diabetic", "Diabetic"))%>%
  filter(PLASMA > 0) %>%
  group_by(PLASMA, RESPONSE) %>%
  summarize(response_total = n())
ggplot(pima_plasma, aes(x=PLASMA, fill=factor(RESPONSE))) +
  geom_density(alpha=0.5) +
  ylab("Density") +
  xlab("Plasma Glucose Concentration in Saliva") +
  ggtitle("Relationship between Diabetic and Non-Diabetic Groups and Plasma Glucose Concentration in Saliva") +
  scale_fill_manual("Diabetic Group", values=c("Red","Blue"))
```


```{r}
#Plot regarding insulin levels
pima_insulin <- pima %>%
  mutate(RESPONSE = ifelse(RESPONSE == 0, "Non-Diabetic", "Diabetic"))%>%
  filter(INSULIN > 0) %>%
  group_by(INSULIN, RESPONSE) %>%
  summarize(response_total = n())
ggplot(pima_insulin, aes(x=INSULIN, fill=factor(RESPONSE))) +
  geom_density(alpha=0.5) +
  ylab("Density") +
  xlab("Two Hours Serum Insulin") +
  ggtitle("Relationship between Diabetic and Non-Diabetic Groups and Two Hours Serum Insulin") +
  scale_fill_manual("Diabetic Group", values=c("Red","Blue"))
```


```{r}
#Plot regarding BMI
pima_body <- pima %>%
  mutate(RESPONSE = ifelse(RESPONSE == 0, "Non-Diabetic", "Diabetic"))%>%
  filter(BODY > 0) %>%
  group_by(BODY, RESPONSE) %>%
  summarize(response_total = n())
ggplot(pima_body, aes(x=BODY, fill=factor(RESPONSE))) +
  geom_density(alpha=0.5) +
  ylab("Density") +
  xlab("Body Mass Index (Weight/Height)") +
  ggtitle("Relationship between Diabetic and Non-Diabetic Groups and Body Mass Index (Weight/Height)") +
  scale_fill_manual("Diabetic Group", values=c("Red","Blue"))
```


```{r}
#Plot regarding pedigree
pima_pedigree <- pima %>%
  mutate(RESPONSE = ifelse(RESPONSE == 0, "Non-Diabetic", "Diabetic"))%>%
  filter(PEDIGREE > 0) %>%
  group_by(PEDIGREE, RESPONSE) %>%
  summarize(response_total = n())
ggplot(pima_pedigree, aes(x=PEDIGREE, fill=factor(RESPONSE))) +
  geom_density(alpha=0.5) +
  ylab("Density") +
  xlab("Pedigree") +
  ggtitle("Relationship between Diabetic and Non-Diabetic Groups and Pedigree") +
<<<<<<< HEAD
  scale_fill_manual(legend_title, values=c("Red","Blue"))
#Pedigree may not be overly important; however, it does seem that there may be a slight relationship.
=======
  scale_fill_manual("Diabetic Group", values=c("Red","Blue"))
#Pedigree may not be overly important; however, it does seem that there may be a slight relationship
>>>>>>> b2096195ddaa531e7ee6b6a130de116f0e038a1f
```


```{r}
#plot regarding number of pregnancy
pima_prg <- pima %>%
  mutate(RESPONSE = ifelse(RESPONSE == 0, "Non-Diabetic", "Diabetic"))%>%
  filter(PRG > 0) %>%
  group_by(PRG, RESPONSE) %>%
  summarize(response_total = n())
ggplot(pima_prg, aes(x=PRG, fill=factor(RESPONSE))) +
  geom_density(alpha=0.5) +
  ylab("Density") +
  xlab("Number of Times Pregnant") +
  ggtitle("Relationship between Diabetic and Non-Diabetic Groups and Number of Times Pregnant") +
  scale_fill_manual("Diabetic Group", values=c("Red","Blue"))
```


## Probability Model of Diabetes in an Individual Pima Female

```{r}
#Checking for an interaction.
fmfull<-glm(RESPONSE~., data=filteredpima)
vif(fmfull)
#Although BODY and THICK are both measures of obesity, it seems that the inclusion of an interaction term is unnecessary.
```


```{r}
filteredmodel <- glm(RESPONSE~., data=filteredpima, family=binomial) %>%
  MASS::stepAIC(trace=FALSE)
summary(filteredmodel) #The stepwise logistic regression returned PLASMA, BODY, and PEDIGREE as significant predictors of type 2 diabetes. AGE is also significant at the 10% significance level.
```


## ROC Curve: An Evaluation of the Probability Model

```{r}
#Creates a ROC Curve and prints out the area under the curve which suggests this model is pretty strong at approximately 86%
predictions <- predict(filteredmodel, data=filteredpima)
plot.roc(filteredpima$RESPONSE, predictions)
auc(filteredpima$RESPONSE, predictions)
```

## Classification Tree

```{r}
#Creates a classifcation tree that denotes the thresholds: defines no diabetes, pre-diabetes, and diabetes
filteredmodelpart <- rpart(RESPONSE ~ ., data = filteredpima)
printcp(filteredmodelpart)
pdf("diabetes.pdf", width = 25, height = 12)
pdf("diabetes.pdf", width = 25, height = 10)
plot(as.party(filteredmodelpart))
dev.off()
diabetes <- mutate(filteredpima, fittedtree = predict(filteredmodelpart))
```


A tree diagram is a way of representing a sequence of events and the purpose of using this diagram is to help predict if a patient has or does not have diabetes depending on certain thresholds of various physical characteristics. In this case, we start at the very top and examine an individual’s plasma glucose concentration and make a move to the left or the right depending on what the individual’s plasma concentration is. Then you would proceed to look at the next physical characteristic and make another move to the left or right again and again until you work your way to the bottom. The boxplots displayed on the bottom generally illustrate 2 sides: a lower risk side on the left and a higher risk side on the right

This diagram shows that plasma glucose concentration is very important because plasma shows up a total of 4 times in the tree so it suggests that plasma level is a pretty big factor in determining if an individual has diabetes.


We also created a classification tree diagram without age and pedigree as age and pedigree are uncontrollable factors.

```{r}
#Removes age and pedigree from the classication tree (see pdf in the folder)
filteredmodelpart <- rpart(RESPONSE ~ . - AGE - PEDIGREE, data = filteredpima)
printcp(filteredmodelpart)
pdf("diabetes1.pdf", width = 25, height = 12)
plot(as.party(filteredmodelpart))
dev.off()
diabetes <- mutate(filteredpima, fittedtree = predict(filteredmodelpart))
```


## Recommendations

We recommend that patients talk to their physician about their treatment choices. Namely focusing on the controllable factors such as plasma glucose concentration in saliva. Diet & exercise are cruical and they help lower levels of the most significant risk factors of diabetes (plasma, BMI, insulin, skinfold thickness, etc.).


